---
title: "CodeBook for Getting and Cleaning Data Course Project"
author: "fjelltronen"
date: "April 14, 2018"
output: 
  html_document:
    keep_md: true
---

### Introduction

This file details the transformations of the UCI HAR Dataset that using the `run_analysis.R` script.

### Data Set

This data set represents data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available at the site where the data was obtained: http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

30 volunteers performed 6 activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone on their waist. A total of 10,299 observations were collected and, for each, 561 attributes were meansured/calculated. 

The data was also randomly split into training (70%) and testing (remaining 30%).

### run_analysis.R

The analysis script (`run_analysis.R`) collects, works with and cleans the data set.

First, it downloads and unzips the data .zip file.

Then, the feature and activity names are read from `features.txt` and `activity_labels.txt`.

```{r}
features <- read.table("./UCI HAR Dataset/features.txt")
head(features)
activities <- read.table("./UCI HAR Dataset/activity_labels.txt",col.names = c("activity_id","activity"))
head(activities)
```

Then, the training data is read from the files in the `train` directory.

```{r}
# use the feature names as column names (changed slightly to be valid column names)
train_features<-read.table(file = "./UCI HAR Dataset/train/X_train.txt",col.names = features$V2)
# 7352 obs. of  561 variables
str(train_features,list.len=6) ## display the first 6 variables only
train_labels<-read.table(file="./UCI HAR Dataset/train/y_train.txt",col.names = c("activity_id"))
train_labels <- merge(train_labels,activities) ## append activity names
# 7352 obs. of  2 variables
head(train_labels)
# 7352 obs. of  1 variable
train_subjects<-read.table(file="./UCI HAR Dataset/train/subject_train.txt",col.names=c("subject"))
str(train_subjects,list.len=6)

# 7352 obs. of  564 variables
train<-cbind(train_labels,train_subjects,train_features)
str(train,list.len=9)
```

Then, the testing data is read from the files in the `test` directory.

```{r}
# use the feature names as column names (changed slightly to be valid column names)
test_features<-read.table(file = "./UCI HAR Dataset/test/X_test.txt",col.names = features$V2)
# 2947 obs. of  561 variables
str(test_features,list.len=6) ## display the first 6 variables only
test_labels<-read.table(file="./UCI HAR Dataset/test/y_test.txt",col.names = c("activity_id"))
test_labels <- merge(test_labels,activities) ## append activity names
# 2947 obs. of  2 variables
head(test_labels)
test_subjects<-read.table(file="./UCI HAR Dataset/test/subject_test.txt",col.names=c("subject"))
# 2947 obs. of  1 variable
str(test_subjects,list.len=6)

test<-cbind(test_labels,test_subjects,test_features)
# 2947 obs. of  564 variables
str(test,list.len=9)
```

Then, the training data and the testing data are combined into a new data frame `allData`.

```{r}
allData<-rbind(train,test)
# 10299 obs. of 564 variables
str(allData,list.len=9)
```

`msData` is computed to have only the measurement columns on the mean and standard deviation from `allData` as well as the `activity` and `subject` columns.

```{r}
# dataset is subset to only certain column names
msData <- subset(allData,select = (grepl("(\\.(mean|std)\\.)|activity|subject",colnames(allData))))
# 10299 obs. of 69 variables
str(msData,list.len=9)
```

From this data frame, the tidy data set is computed to have the average of each variable for each activity and each subject. The dataset is printed to the `gcd-tidy-data.txt` file.

```{r}
avgData <- aggregate(msData[,-c(1:3)], msData[,c(1:3)], mean)
# 40 obs. of 69 variables
head(avgData[,1:9])
```

